Accelerating Neural Network Training using Arbitrary Precision Approximating Matrix Multiplication Algorithms

### 2021-07-06 latex/ICPP21/paper.tex
converted input of tikz files to including pdf versions
add CCS concepts and keywords, converted away from cleveref

% cleveref
\usepackage{hyperref}
\usepackage[nameinlink]{cleveref}
\crefname{section}{\S}{Sections}
\crefname{figure}{Fig.}{Figs.}
\Crefname{figure}{Figure}{Figures}
\crefname{table}{Table}{Tables}

% tikz
\usepackage{tikz,pgfplots}
\usetikzlibrary{positioning}
\usetikzlibrary{external} 
\tikzexternalize[optimize=false, prefix=tikz/]

@@

@@

One of the reasons such algorithms are impractical is that they are based on so-called \emph{Arbitrary Precision Approximating} (APA) algorithms, which we describe in detail in \cref{sec:APA}.
In exact arithmetic, these algorithms compute an approximation of the correct result, where the error is polynomial in a nonzero parameter of the algorithm \cite{BLR80}.
That is, in exact arithmetic, the error can be made arbitrarily small.
In floating point arithmetic, however, there is a lower bound on the approximation error that depends on the working precision and properties of the algorithm.
Thus, APA algorithms are often considered to have insufficient accuracy for most applications and have largely been overlooked as practical tools despite their performance potential and ability to outperform exact algorithms \cite{BB15}.

@@

A key characteristic of matrix multiplication, first demonstrated by Bini et al.~\cite{BCRL79}, is that it can be approximated to arbitrary accuracy with less computation than required by exact algorithms.
An APA algorithm is one that takes as input $A$, $B$, and a scalar parameter $0<\lambda<1$ and computes 
\begin{equation}
\label{eq:APAapprox}
\hat C = A\cdot B + \lambda E + O(\lambda^2),
\end{equation}
where $E$ is an error matrix that depends on $A$ and $B$ and corresponds to the leading term of the error polynomial.
Thus, in exact arithmetic, letting $\lambda \rightarrow 0$ achieves arbitrarily small approximation error.
In floating point arithmetic, choosing too small a value for $\lambda$ leads to accumulation of roundoff error that exceeds the approximation error.
We discuss optimizing $\lambda$ in \cref{sec:APAerr}.

@@

For a concrete example, we reproduce the rule for the algorithm developed by Bini et al.~\cite{BCRL79} for the $\dims{3,2,2}$ case (multiplying a $3\times 2$ matrix $A$ by a $2\times 2$ matrix $B$), where we use the following notation for input and output matrices:
$$
\begin{bmatrix} A_{11} & A_{12} \\ A_{21} & A_{22} \\ A_{31} & A_{32}  \end{bmatrix} \cdot
\begin{bmatrix} B_{11} & B_{12} \\ B_{21} & B_{22} \end{bmatrix}=
\begin{bmatrix} C_{11} & C_{12} \\ C_{21} & C_{22} \\ C_{31} & C_{32} \end{bmatrix} .$$
Bini's rule is given by
\begin{align*}
M_1 =& (A_{11} + A_{22}) \cdot (\lambda B_{11} + B_{22}) \\
M_2 =& A_{22}\cdot  (-B_{21} - B_{22}) \\
M_3 =& A_{11}\cdot B_{22} \\
M_4 =& (\lambda A_{12} + A_{22})\cdot (-\lambda B_{11} + B_{21}) \\
M_5 =& (A_{11} + \lambda A_{12}) \cdot (\lambda B_{12} + B_{22}) \\
M_6 =& (A_{21} + A_{32}) \cdot (B_{11} + \lambda B_{22}) \\
M_7 =& A_{21} \cdot (-B_{11} - B_{12}) \\
M_8 =& A_{32} \cdot B_{11}\\
M_9 =& (A_{21} + \lambda A_{31}) \cdot (B_{12} - \lambda B_{22})\\
M_{10} =& (\lambda A_{31} + A_{32}) \cdot (B_{12} - \lambda B_{22})\\
\end{align*}
\begin{align*}
\hat C_{11} =& \lambda^{-1} (M_1 + M_2 - M_3 + M_4) \\
\hat C_{12} =& \lambda^{-1} (-M_3 + M_5) \\
\hat C_{21} =& M_4 + M_6 - M_{10}\\
\hat C_{22} =& M_1 - M_5 + M_9\\
\hat C_{31} =& \lambda^{-1} (-M_8 + M_{10}) \\
\hat C_{32} =& \lambda^{-1} (M_6 + M_7 - M_8 + M_9).\\
\end{align*}
For example, this rule computes the first entry of the output matrix as $\hat C_{11} = A_{11}  B_{11} + A_{12}  B_{21} - \lambda A_{12}  B_{11}$.
Thus, in this case, the first entry of the error matrix $E$ in \cref{eq:APAapprox} is $E_{11} = A_{12}  B_{11}$.

@@

The two parameters of the APA algorithm specify the contribution of the approximation and roundoff errors, respectively.
The first parameter, $\sigma$, is the smallest positive exponent of the error polynomial and represents the approximation error.
\Cref{eq:APAapprox} shows the error as a polynomial of $\lambda$ whose leading term is linear in $\lambda$.
If an algorithm satisfies \cref{eq:APAapprox} with $E\neq 0$, then $\sigma=1$.
However, if $E=0$, then $\sigma>1$ is the degree of the leading monomial.
Larger $\sigma$ implies smaller error due to the algorithm, though all of the APA algorithms we consider have $\sigma=1$.

The second parameter, $\varphi$, is the largest (in absolute value) negative exponent of the algorithm, computed as the largest sum of negative exponents across all triplets of matrices.
This parameter represents the effect of roundoff error caused by floating point arithmetic involving the largest intermediate values computed by the algorithm.
For example, the triplet given in \cref{eq:triplet} yields a sum of negative exponents of $0+0+1=1$, and in the case of Bini's algorithm, no other triplet has a larger sum, so $\varphi=1$ for that algorithm.
Smaller $\varphi$ implies smaller error due to roundoff, and the APA algorithms we consider exhibit a range of values.

Given these two contributions to the numerical error, $\lambda$ can be optimized to balance the effects based on parameters $\sigma$ and $\varphi$ (and $d$).
As shown by Bini, Lotti, and Romani \cite{BLR80}, the optimal $\lambda$ should be set to $\Theta(2^{-d/(\sigma+\varphi)})$.
Using this value of $\lambda$, the numerical error incurred by the algorithm will be bounded by $O(2^{-d\sigma/(\sigma+\varphi)})$.
Taking Bini's algorithm as an example, we have $\sigma=\varphi=1$, so the error is $O(2^{-d/2})$, or the square root of working precision.
Note that if multiple recursive steps are used, then $\varphi$ increases proportional to the number of steps, so straightforward optimization of $\lambda$ results in error of $O(2^{-d\sigma/(\sigma+s\varphi)})$ for $s$ recursive steps.
The parameters and minimum error for the algorithms we consider are presented in \cref{sec:APAprops}.

We show empirical error results for uniform random inputs of varying dimension in \cref{fig:accuracy}, as compared to the classical algorithm.
We measure the relative Frobenius norm error, or $\|C-\hat C\|_F/\|C\|_F$, where $\hat C$ is computed by each algorithm and $C$ is computed using the classical algorithm in double precision.
In order to choose the optimal $\lambda$ value for each algorithm, we tested the 5 powers of 2 closest to the theoretical optimal value and chose the best.

\begin{figure}
\centering
\renewcommand{\datafile}{data/matmul_acc.dat}
\input{plots/accuracy}
\caption{Relative Frobenius norm error for APA algorithms on random inputs.}
\label{fig:accuracy}
\end{figure}

Overall, we see little fluctuation of the error over matrix dimension, and the theoretical error bound is an upper bound on all empirical errors.
Note that the legend is ordered according to the error parameters, and the empirical error generally follows this ordering.
The two most accurate APA algorithms are the least recently discovered: Bini's $\dims{3,2,2}$ and Schonhage's $\dims{3,3,3;21}$ (note that we indicate the rank of the algorithm here to distinguish from the alternative rank-20 algorithm with the same dimensions).
Algorithms that offer more potential speedup tend to be less accurate.
Two algorithms with smaller error than expected are $\dims{5,5,5}$ and $\dims{7,2,2}$, which is explained by the coefficients of those algorithms including fractional pre-factors.
While $\phi$ is computed using the exponents of the largest intermediate term of $\lambda$, the leading term of $\dims{5,5,5}$ has a constant of $1/4$, lessening its magnitude. 
We study the effects of the matrix multiplication error on neural network training and test accuracy in \cref{sec:NNacc}.

@@

\Cref{tab:algs} shows the key performance and accuracy properties of the APA algorithms we consider.
Each row corresponds to an algorithm, and the first row includes the classical algorithm for comparison.
For all algorithms, we assume only 1 recursive step is used, though the speedup and error for more steps can be readily calculated from the parameters.
The first column gives the reference where the algorithm was first specified.
The second block column demonstrates the possible performance improvement, where speedup is calculated as $(mnk/r - 1)\cdot100\%$ for $\dims{m,n,k}$ and rank $r$.
The third block column shows error parameters, and the error is calculated as $2^{-d\cdot\sigma/(\sigma+\varphi)}$ with $d=23$, corresponding to single precision.

@@

\begin{figure}
\centering
\input{figs/hybrid}
\caption{Illustration of hybrid parallelization strategy for $r=10$ and 4 threads.  Each thread is assigned two multiplications to compute using single-threaded \texttt{gemm} and the two remaining multiplications are performed using multithreaded \texttt{gemm}.}
\label{fig:hybrid}
\end{figure}

\begin{figure*}[t!]
%
\subfloat[One thread]{
\centering
\renewcommand{\datafile}{data/matmul_seq.dat}
\renewcommand{\threads}{1}
\input{plots/matmul_perf}
%\caption{Single-threaded matrix multiplication performance}
\label{fig:matmul_seq}}
%
\hfill
\subfloat[Six threads]{
\centering
\renewcommand{\datafile}{data/matmul_par_6.dat}
\renewcommand{\threads}{6}
\input{plots/matmul_perf}
%\caption{Multithreaded matrix multiplication performance (6 threads)}
\label{fig:matmul_par_6}}
%
\hfill
\subfloat[Twelve threads]{
\centering
\renewcommand{\datafile}{data/matmul_par_12.dat}
\renewcommand{\threads}{12}
\input{plots/matmul_perf}
%\caption{Multithreaded matrix multiplication performance (6 threads)}
\label{fig:matmul_par_12}}
%
\caption{Single- and multi-threaded square matrix multiplication performance, relative to $2n^3$ operations.  The machine peak with respect to classical algorithms is given as a dotted line.}
\label{fig:matmul_perf}
\end{figure*}

@@

As described in \cref{sec:APAalgs}, each algorithm can be encoded by a set of triplets of coefficient matrices.
From this representation, we generate recursive code that computes the linear combinations for the inputs to each multiplication and combines the outputs.
The linear combinations are computed using a ``write-once'' strategy that was found to be most efficient in terms of memory bandwidth and performance.
The multiplications are either computed recursively or via a call to \texttt{gemm}, the interface to a highly efficient BLAS implementation of the classical algorithm.
Because we use only 1 recursive step in our experiments, there are no recursive calls, and every multiplication is performed by \texttt{gemm}.
We note that the original code generation tool was designed primarily for exact algorithms, and we generalized it to apply to all of the APA algorithms we consider here, some of which have more complicated coefficients than previously considered.

We adopt the ``hybrid'' parallelization strategy proposed by Benson and Ballard \cite{BB15}.
All linear combinations are parallelized in a straightforward way, in order to maximize the memory bandwidth of the machine.
The multiplications are parallelized as follows: given $r$ multiplications and $p$ threads, with $r=p\cdot q + \ell$ for integers $q$ and $\ell<p$, we assign each thread $q$ multiplications to be performed independently, and the remaining $\ell$ multiplications are performed by all threads using the multithreaded implementation of \texttt{gemm}.
\Cref{fig:hybrid} shows the hybrid strategy for $r=10$ (Bini's algorithm) and $p=4$.

@@

\Cref{fig:matmul_seq} reports the sequential performance of the APA algorithms in comparison to the most efficient implementation of the classical algorithm, MKL's single-precision \texttt{gemm}.
The y-axis is the \emph{effective} GFLOPS, which is measured as $\num{1e-9} \cdot 2n^3 / \text{ time}$.
That is, the GFLOPS reported for APA algorithms is not true performance, as they perform fewer flops than the classical algorithm.
We use this metric to be able to compare relative times across algorithms performing different amounts of computation.
The machine peak for a classical algorithm is given by the horizontal dotted line.

@@

We report parallel performance in \cref{fig:matmul_par_6,fig:matmul_par_12} for running with 6 threads (one socket) and 12 threads (both sockets).
Overall, speedups of APA algorithms over classical are reduced in the parallel case.
Again, the theoretical speedup is based on a reduction in the multiplications, and the overhead of additions is the biggest impediment to realizing that speedup.
In the parallel case, the additions can become an even larger bottleneck because the additions are memory bandwidth bound, and the memory bandwidth does not scale with the number of cores \cite{BB15}.
While we can expect close to linear speedup with cores on the multiplications (which are perfectly load balanced), linear scaling is impossible to achieve with the additions.
We also note that when the hybrid method uses all threads on remainder multiplications, the smaller dimensions make it harder for multithreaded \texttt{gemm} to scale as well as it can on the original matrix dimensions.

In the performance results for 6 threads (\cref{fig:matmul_par_6}), we see that many of the algorithms start to outperform classical around dimension 2000, though some poor performance is observed for algorithms and particular matrix dimensions.
The fastest algorithms (e.g., $\dims{4,4,4}$ and $\dims{4,4,2}$) achieve a speedup of up to 25\% over \texttt{gemm} and exceed the machine peak for classical algorithms.
We note that the $\dims{4,4,2}$ algorithm has 24 subproblems, which is a multiple of 6, so there are no remainder subproblems that require all threads.

\Cref{fig:matmul_par_12} shows the results for 12 threads.
Here we see a majority of the algorithms are slower than the classical algorithm, even for large matrices.
We attribute the poor performance to the effect of lower \texttt{gemm} performance for smaller matrices, as well as a lack of NUMA-aware optimization.
The ``ramp-up'' range of \texttt{gemm} performance is much shallower for 12 threads than for 6 threads, not achieving the plateau performance until dimension 4000 or so.
This implies that the dimension of the submultiplications does not fall on the plateau for any APA algorithm, so the remainder multiplications suffer from poor parallel performance.
The algorithm with no remainder multiplications, $\dims{4,2,2}$, does not suffer this problem and maintains higher performance.
It exceeds \texttt{gemm}'s effective performance at dimension 4000, exceeds the peak parallel performance of any classical algorithm for larger dimensions, and achieves a 21\% speedup over \texttt{gemm} at dimension 8192 for an effective rate of 389 GFLOPS.

@@

\begin{figure*}[ht!]
%
\subfloat[Training accuracy]{
\centering
\renewcommand{\datafile}{data/NN_acc.dat}
\renewcommand{\set}{train}
\input{plots/NN_acc}
\label{fig:NN_acc_train}}
%
\hfill
\subfloat[Test accuracy]{
\centering
\renewcommand{\datafile}{data/NN_acc.dat}
\renewcommand{\set}{test}
\input{plots/NN_acc}
\label{fig:NN_acc_test}}
%
\caption{MLP network accuracy for MNIST data set}
\label{fig:NN_acc}
\end{figure*}

\begin{figure}[hb!]
\centering
\input{figs/neuralnet}
\caption{Multi-Layer Perceptron network structure}
\label{fig:neuralnet}
\end{figure}

@@

To illustrate the effectiveness of APA algorithms in neural network training, we consider Multi-Layer Perceptron networks, as shown in \cref{fig:neuralnet}.
We use TensorFlow 2.2.0 to build the networks and implement custom operators for matrix multiplication.
For internal layers, we use the custom operator for both forward propagation and gradient calculation.
For fair comparison against the classical algorithm, we use a custom classical operator that directly calls \texttt{gemm}, which significantly outperformed TensorFlow's built-in matrix multiplication operator for fully connected layers.

We show an MLP network with two hidden layers in \cref{fig:neuralnet}.
The dimensions in the figure match those of the accuracy experiment described in \cref{sec:NNacc}.
For the performance experiment of \cref{sec:NNperf}, we use an MLP with four hidden layers and varying numbers of nodes in the hidden layers.

@@

\Cref{fig:NN_acc_train} plots the accuracy on the training data over each epoch, and \cref{fig:NN_acc_test} shows the test accuracy over epochs. 
The matrix multiplication error introduced from these algorithms has minimal effect on the training error; all algorithms yield convergence to nearly full accuracy after 20 epochs or so.
We see more variability in the test error due to the sensitivity inherent in the generalizability of the model, though all algorithms achieve between 97\% and 99\% test accuracy.

@@

\begin{figure*}[t!]
%
\subfloat[One thread]{
\centering
\renewcommand{\datafile}{data/NN_seq.dat}
\renewcommand{\threads}{1}
\input{plots/NN_reltime}
\label{fig:NNreltime_seq}}
%
\hfill
\subfloat[Six threads]{
\centering
\renewcommand{\datafile}{data/NN_par_6.dat}
\renewcommand{\threads}{6}
\input{plots/NN_reltime}
\label{fig:NNreltime_par_6}}
%
\hfill
\subfloat[Twelve threads]{
\centering
\renewcommand{\datafile}{data/NN_par_12.dat}
\renewcommand{\threads}{12}
\input{plots/NN_reltime}
\label{fig:NNreltime_par_12}}
%
\caption{Network training time relative to using classical matrix multiplication}
\label{fig:NNreltime}
\end{figure*}

>>>>

% cleveref
\usepackage{hyperref}
%\usepackage[nameinlink]{cleveref}
%\crefname{section}{\S}{Sections}
%\crefname{figure}{Fig.}{Figs.}
%\Crefname{figure}{Figure}{Figures}
%\crefname{table}{Table}{Tables}

% tikz
%\usepackage{tikz,pgfplots}
%\usetikzlibrary{positioning}
%\usetikzlibrary{external} 
%\tikzexternalize[optimize=false, prefix=tikz/]

@@

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10002950.10003705.10011686</concept_id>
       <concept_desc>Mathematics of computing~Mathematical software performance</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
  <concept>
       <concept_id>10010147.10010257.10010293.10010294</concept_id>
       <concept_desc>Computing methodologies~Neural networks</concept_desc>
       <concept_significance>500</concept_significance>
       </concept>
 </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Mathematics of computing~Mathematical software performance}
\ccsdesc[500]{Computing methodologies~Neural networks}

\keywords{neural networks, multilayer perceptrons, APA matrix multiplication}

@@

One of the reasons such algorithms are impractical is that they are based on so-called \emph{Arbitrary Precision Approximating} (APA) algorithms, which we describe in detail in \S~\ref{sec:APA}.
In exact arithmetic, these algorithms compute an approximation of the correct result, where the error is polynomial in a nonzero parameter of the algorithm \cite{BLR80}.
That is, in exact arithmetic, the error can be made arbitrarily small.
In floating point arithmetic, however, there is a lower bound on the approximation error that depends on the working precision and properties of the algorithm.
Thus, APA algorithms are often considered to have insufficient accuracy for most applications and have largely been overlooked as practical tools despite their performance potential and ability to outperform exact algorithms \cite{BB15}.

@@

A key characteristic of matrix multiplication, first demonstrated by Bini et al.~\cite{BCRL79}, is that it can be approximated to arbitrary accuracy with less computation than required by exact algorithms.
An APA algorithm is one that takes as input $A$, $B$, and a scalar parameter $0<\lambda<1$ and computes 
\begin{equation}
\label{eq:APAapprox}
\hat C = A\cdot B + \lambda E + O(\lambda^2),
\end{equation}
where $E$ is an error matrix that depends on $A$ and $B$ and corresponds to the leading term of the error polynomial.
Thus, in exact arithmetic, letting $\lambda \rightarrow 0$ achieves arbitrarily small approximation error.
In floating point arithmetic, choosing too small a value for $\lambda$ leads to accumulation of roundoff error that exceeds the approximation error.
We discuss optimizing $\lambda$ in \S~\ref{sec:APAerr}.

@@

For a concrete example, we reproduce the rule for the algorithm developed by Bini et al.~\cite{BCRL79} for the $\dims{3,2,2}$ case (multiplying a $3\times 2$ matrix $A$ by a $2\times 2$ matrix $B$), where we use the following notation for input and output matrices:
$$
\begin{bmatrix} A_{11} & A_{12} \\ A_{21} & A_{22} \\ A_{31} & A_{32}  \end{bmatrix} \cdot
\begin{bmatrix} B_{11} & B_{12} \\ B_{21} & B_{22} \end{bmatrix}=
\begin{bmatrix} C_{11} & C_{12} \\ C_{21} & C_{22} \\ C_{31} & C_{32} \end{bmatrix} .$$
Bini's rule is given by
\begin{align*}
M_1 =& (A_{11} + A_{22}) \cdot (\lambda B_{11} + B_{22}) \\
M_2 =& A_{22}\cdot  (-B_{21} - B_{22}) \\
M_3 =& A_{11}\cdot B_{22} \\
M_4 =& (\lambda A_{12} + A_{22})\cdot (-\lambda B_{11} + B_{21}) \\
M_5 =& (A_{11} + \lambda A_{12}) \cdot (\lambda B_{12} + B_{22}) \\
M_6 =& (A_{21} + A_{32}) \cdot (B_{11} + \lambda B_{22}) \\
M_7 =& A_{21} \cdot (-B_{11} - B_{12}) \\
M_8 =& A_{32} \cdot B_{11}\\
M_9 =& (A_{21} + \lambda A_{31}) \cdot (B_{12} - \lambda B_{22})\\
M_{10} =& (\lambda A_{31} + A_{32}) \cdot (B_{12} - \lambda B_{22})\\
\end{align*}
\begin{align*}
\hat C_{11} =& \lambda^{-1} (M_1 + M_2 - M_3 + M_4) \\
\hat C_{12} =& \lambda^{-1} (-M_3 + M_5) \\
\hat C_{21} =& M_4 + M_6 - M_{10}\\
\hat C_{22} =& M_1 - M_5 + M_9\\
\hat C_{31} =& \lambda^{-1} (-M_8 + M_{10}) \\
\hat C_{32} =& \lambda^{-1} (M_6 + M_7 - M_8 + M_9).\\
\end{align*}
For example, this rule computes the first entry of the output matrix as $\hat C_{11} = A_{11}  B_{11} + A_{12}  B_{21} - \lambda A_{12}  B_{11}$.
Thus, in this case, the first entry of the error matrix $E$ in eq.~\eqref{eq:APAapprox} is $E_{11} = A_{12}  B_{11}$.

@@

The two parameters of the APA algorithm specify the contribution of the approximation and roundoff errors, respectively.
The first parameter, $\sigma$, is the smallest positive exponent of the error polynomial and represents the approximation error.
Equation~\eqref{eq:APAapprox} shows the error as a polynomial of $\lambda$ whose leading term is linear in $\lambda$.
If an algorithm satisfies eq.~\eqref{eq:APAapprox} with $E\neq 0$, then $\sigma=1$.
However, if $E=0$, then $\sigma>1$ is the degree of the leading monomial.
Larger $\sigma$ implies smaller error due to the algorithm, though all of the APA algorithms we consider have $\sigma=1$.

The second parameter, $\varphi$, is the largest (in absolute value) negative exponent of the algorithm, computed as the largest sum of negative exponents across all triplets of matrices.
This parameter represents the effect of roundoff error caused by floating point arithmetic involving the largest intermediate values computed by the algorithm.
For example, the triplet given in eq.~\eqref{eq:triplet} yields a sum of negative exponents of $0+0+1=1$, and in the case of Bini's algorithm, no other triplet has a larger sum, so $\varphi=1$ for that algorithm.
Smaller $\varphi$ implies smaller error due to roundoff, and the APA algorithms we consider exhibit a range of values.

Given these two contributions to the numerical error, $\lambda$ can be optimized to balance the effects based on parameters $\sigma$ and $\varphi$ (and $d$).
As shown by Bini, Lotti, and Romani \cite{BLR80}, the optimal $\lambda$ should be set to $\Theta(2^{-d/(\sigma+\varphi)})$.
Using this value of $\lambda$, the numerical error incurred by the algorithm will be bounded by $O(2^{-d\sigma/(\sigma+\varphi)})$.
Taking Bini's algorithm as an example, we have $\sigma=\varphi=1$, so the error is $O(2^{-d/2})$, or the square root of working precision.
Note that if multiple recursive steps are used, then $\varphi$ increases proportional to the number of steps, so straightforward optimization of $\lambda$ results in error of $O(2^{-d\sigma/(\sigma+s\varphi)})$ for $s$ recursive steps.
The parameters and minimum error for the algorithms we consider are presented in \S~\ref{sec:APAprops}.

We show empirical error results for uniform random inputs of varying dimension in Fig.~\ref{fig:accuracy}, as compared to the classical algorithm.
We measure the relative Frobenius norm error, or $\|C-\hat C\|_F/\|C\|_F$, where $\hat C$ is computed by each algorithm and $C$ is computed using the classical algorithm in double precision.
In order to choose the optimal $\lambda$ value for each algorithm, we tested the 5 powers of 2 closest to the theoretical optimal value and chose the best.

\begin{figure}
\centering
\renewcommand{\datafile}{data/matmul_acc.dat}
%\input{plots/accuracy}
\includegraphics{tikz-taps/paper-figure0}
\caption{Relative Frobenius norm error for APA algorithms on random inputs.}
\label{fig:accuracy}
\end{figure}

Overall, we see little fluctuation of the error over matrix dimension, and the theoretical error bound is an upper bound on all empirical errors.
Note that the legend is ordered according to the error parameters, and the empirical error generally follows this ordering.
The two most accurate APA algorithms are the least recently discovered: Bini's $\dims{3,2,2}$ and Schonhage's $\dims{3,3,3;21}$ (note that we indicate the rank of the algorithm here to distinguish from the alternative rank-20 algorithm with the same dimensions).
Algorithms that offer more potential speedup tend to be less accurate.
Two algorithms with smaller error than expected are $\dims{5,5,5}$ and $\dims{7,2,2}$, which is explained by the coefficients of those algorithms including fractional pre-factors.
While $\phi$ is computed using the exponents of the largest intermediate term of $\lambda$, the leading term of $\dims{5,5,5}$ has a constant of $1/4$, lessening its magnitude. 
We study the effects of the matrix multiplication error on neural network training and test accuracy in \S~\ref{sec:NNacc}.

@@

Table~\ref{tab:algs} shows the key performance and accuracy properties of the APA algorithms we consider.
Each row corresponds to an algorithm, and the first row includes the classical algorithm for comparison.
For all algorithms, we assume only 1 recursive step is used, though the speedup and error for more steps can be readily calculated from the parameters.
The first column gives the reference where the algorithm was first specified.
The second block column demonstrates the possible performance improvement, where speedup is calculated as $(mnk/r - 1)\cdot100\%$ for $\dims{m,n,k}$ and rank $r$.
The third block column shows error parameters, and the error is calculated as $2^{-d\cdot\sigma/(\sigma+\varphi)}$ with $d=23$, corresponding to single precision.

@@

\begin{figure}
\centering
%\input{figs/hybrid}
\includegraphics{tikz-taps/paper-figure1.pdf}
\caption{Illustration of hybrid parallelization strategy for $r=10$ and 4 threads.  Each thread is assigned two multiplications to compute using single-threaded \texttt{gemm} and the two remaining multiplications are performed using multithreaded \texttt{gemm}.}
\label{fig:hybrid}
\end{figure}

\begin{figure*}[t!]
%
\subfloat[One thread]{
\centering
\renewcommand{\datafile}{data/matmul_seq.dat}
\renewcommand{\threads}{1}
%\input{plots/matmul_perf}
\includegraphics{tikz-taps/paper-figure2.pdf}
%\caption{Single-threaded matrix multiplication performance}
\label{fig:matmul_seq}}
%
\hfill
\subfloat[Six threads]{
\centering
\renewcommand{\datafile}{data/matmul_par_6.dat}
\renewcommand{\threads}{6}
%\input{plots/matmul_perf}
\includegraphics{tikz-taps/paper-figure3.pdf}
%\caption{Multithreaded matrix multiplication performance (6 threads)}
\label{fig:matmul_par_6}}
%
\hfill
\subfloat[Twelve threads]{
\centering
\renewcommand{\datafile}{data/matmul_par_12.dat}
\renewcommand{\threads}{12}
%\input{plots/matmul_perf}
\includegraphics{tikz-taps/paper-figure4.pdf}
%\caption{Multithreaded matrix multiplication performance (6 threads)}
\label{fig:matmul_par_12}}
%
\caption{Single- and multi-threaded square matrix multiplication performance, relative to $2n^3$ operations.  The machine peak with respect to classical algorithms is given as a dotted line.}
\label{fig:matmul_perf}
\end{figure*}

@@

As described in \S~\ref{sec:APAalgs}, each algorithm can be encoded by a set of triplets of coefficient matrices.
From this representation, we generate recursive code that computes the linear combinations for the inputs to each multiplication and combines the outputs.
The linear combinations are computed using a ``write-once'' strategy that was found to be most efficient in terms of memory bandwidth and performance.
The multiplications are either computed recursively or via a call to \texttt{gemm}, the interface to a highly efficient BLAS implementation of the classical algorithm.
Because we use only 1 recursive step in our experiments, there are no recursive calls, and every multiplication is performed by \texttt{gemm}.
We note that the original code generation tool was designed primarily for exact algorithms, and we generalized it to apply to all of the APA algorithms we consider here, some of which have more complicated coefficients than previously considered.

We adopt the ``hybrid'' parallelization strategy proposed by Benson and Ballard \cite{BB15}.
All linear combinations are parallelized in a straightforward way, in order to maximize the memory bandwidth of the machine.
The multiplications are parallelized as follows: given $r$ multiplications and $p$ threads, with $r=p\cdot q + \ell$ for integers $q$ and $\ell<p$, we assign each thread $q$ multiplications to be performed independently, and the remaining $\ell$ multiplications are performed by all threads using the multithreaded implementation of \texttt{gemm}.
Figure~\ref{fig:hybrid} shows the hybrid strategy for $r=10$ (Bini's algorithm) and $p=4$.

@@

Figure~\ref{fig:matmul_seq} reports the sequential performance of the APA algorithms in comparison to the most efficient implementation of the classical algorithm, MKL's single-precision \texttt{gemm}.
The y-axis is the \emph{effective} GFLOPS, which is measured as $\num{1e-9} \cdot 2n^3 / \text{ time}$.
That is, the GFLOPS reported for APA algorithms is not true performance, as they perform fewer flops than the classical algorithm.
We use this metric to be able to compare relative times across algorithms performing different amounts of computation.
The machine peak for a classical algorithm is given by the horizontal dotted line.

@@

We report parallel performance in Figs.~\ref{fig:matmul_par_6} and \ref{fig:matmul_par_12} for running with 6 threads (one socket) and 12 threads (both sockets).
Overall, speedups of APA algorithms over classical are reduced in the parallel case.
Again, the theoretical speedup is based on a reduction in the multiplications, and the overhead of additions is the biggest impediment to realizing that speedup.
In the parallel case, the additions can become an even larger bottleneck because the additions are memory bandwidth bound, and the memory bandwidth does not scale with the number of cores \cite{BB15}.
While we can expect close to linear speedup with cores on the multiplications (which are perfectly load balanced), linear scaling is impossible to achieve with the additions.
We also note that when the hybrid method uses all threads on remainder multiplications, the smaller dimensions make it harder for multithreaded \texttt{gemm} to scale as well as it can on the original matrix dimensions.

In the performance results for 6 threads (Fig.~\ref{fig:matmul_par_6}), we see that many of the algorithms start to outperform classical around dimension 2000, though some poor performance is observed for algorithms and particular matrix dimensions.
The fastest algorithms (e.g., $\dims{4,4,4}$ and $\dims{4,4,2}$) achieve a speedup of up to 25\% over \texttt{gemm} and exceed the machine peak for classical algorithms.
We note that the $\dims{4,4,2}$ algorithm has 24 subproblems, which is a multiple of 6, so there are no remainder subproblems that require all threads.

Figure~\ref{fig:matmul_par_12} shows the results for 12 threads.
Here we see a majority of the algorithms are slower than the classical algorithm, even for large matrices.
We attribute the poor performance to the effect of lower \texttt{gemm} performance for smaller matrices, as well as a lack of NUMA-aware optimization.
The ``ramp-up'' range of \texttt{gemm} performance is much shallower for 12 threads than for 6 threads, not achieving the plateau performance until dimension 4000 or so.
This implies that the dimension of the submultiplications does not fall on the plateau for any APA algorithm, so the remainder multiplications suffer from poor parallel performance.
The algorithm with no remainder multiplications, $\dims{4,2,2}$, does not suffer this problem and maintains higher performance.
It exceeds \texttt{gemm}'s effective performance at dimension 4000, exceeds the peak parallel performance of any classical algorithm for larger dimensions, and achieves a 21\% speedup over \texttt{gemm} at dimension 8192 for an effective rate of 389 GFLOPS.

@@

\begin{figure*}[ht!]
%
\subfloat[Training accuracy]{
\centering
\renewcommand{\datafile}{data/NN_acc.dat}
\renewcommand{\set}{train}
%\input{plots/NN_acc}
\includegraphics{tikz-taps/paper-figure5.pdf}
\label{fig:NN_acc_train}}
%
\hfill
\subfloat[Test accuracy]{
\centering
\renewcommand{\datafile}{data/NN_acc.dat}
\renewcommand{\set}{test}
%\input{plots/NN_acc}
\includegraphics{tikz-taps/paper-figure6.pdf}
\label{fig:NN_acc_test}}
%
\caption{MLP network accuracy for MNIST data set}
\label{fig:NN_acc}
\end{figure*}

\begin{figure}[hb!]
\centering
%\input{figs/neuralnet}
\includegraphics{tikz-taps/paper-figure7.pdf}
\caption{Multi-Layer Perceptron network structure}
\label{fig:neuralnet}
\end{figure}

@@

To illustrate the effectiveness of APA algorithms in neural network training, we consider Multi-Layer Perceptron networks, as shown in Fig.~\ref{fig:neuralnet}.
We use TensorFlow 2.2.0 to build the networks and implement custom operators for matrix multiplication.
For internal layers, we use the custom operator for both forward propagation and gradient calculation.
For fair comparison against the classical algorithm, we use a custom classical operator that directly calls \texttt{gemm}, which significantly outperformed TensorFlow's built-in matrix multiplication operator for fully connected layers.

We show an MLP network with two hidden layers in Fig.~\ref{fig:neuralnet}.
The dimensions in the figure match those of the accuracy experiment described in \S~\ref{sec:NNacc}.
For the performance experiment of \S~\ref{sec:NNperf}, we use an MLP with four hidden layers and varying numbers of nodes in the hidden layers.

@@

Figure~\ref{fig:NN_acc_train} plots the accuracy on the training data over each epoch, and Fig.~\ref{fig:NN_acc_test} shows the test accuracy over epochs. 
The matrix multiplication error introduced from these algorithms has minimal effect on the training error; all algorithms yield convergence to nearly full accuracy after 20 epochs or so.
We see more variability in the test error due to the sensitivity inherent in the generalizability of the model, though all algorithms achieve between 97\% and 99\% test accuracy.

@@

\begin{figure*}[t!]
%
\subfloat[One thread]{
\centering
\renewcommand{\datafile}{data/NN_seq.dat}
\renewcommand{\threads}{1}
%\input{plots/NN_reltime}
\includegraphics{tikz-taps/paper-figure8.pdf}
\label{fig:NNreltime_seq}}
%
\hfill
\subfloat[Six threads]{
\centering
\renewcommand{\datafile}{data/NN_par_6.dat}
\renewcommand{\threads}{6}
%\input{plots/NN_reltime}
\includegraphics{tikz-taps/paper-figure9.pdf}
\label{fig:NNreltime_par_6}}
%
\hfill
\subfloat[Twelve threads]{
\centering
\renewcommand{\datafile}{data/NN_par_12.dat}
\renewcommand{\threads}{12}
%\input{plots/NN_reltime}
\includegraphics{tikz-taps/paper-figure10.pdf}
\label{fig:NNreltime_par_12}}
%
\caption{Network training time relative to using classical matrix multiplication}
\label{fig:NNreltime}
\end{figure*}
